---
title: 'Reproducible Research: Peer Assessment 2'
author: "Ankit Singhal"
date: "14 September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Explore the NOAA Storm Database

### About the Report
In this report, we aim to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Basic settings

```{r Library_Loading}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
```

### Data Loading and Processing
First, we download the data file and unzip it.
```{r File_Downloading}
setwd("C:\\Machine Learning\\ReepData_Project2")

if (!"stormData.csv.bz2" %in% dir("./data/")) {
   url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url, "StormData.csv.bz2")
    bunzip2("StormData.csv.bz2", "StormData.csv")
}


```  
Reading csv file.  
```{r Reading_File}
if (!"df" %in% ls()) {
    df <- read.csv("data/stormData.csv")
}
```  
##Data Processing  

###Health Impact  

To evaluate the health impact, the total fatalities and the total injuries for each event type (EVTYPE) are calculated. The codes for this calculation are shown as follows.  
```{r HealthImpact}
library(dplyr)
df.fatalities <- df %>% select(EVTYPE, FATALITIES) %>% group_by(EVTYPE) %>% summarise(total.fatalities = sum(FATALITIES)) %>% arrange(-total.fatalities)
head(df.fatalities, 10)

df.injuries <- df %>% select(EVTYPE, INJURIES) %>% group_by(EVTYPE) %>% summarise(total.injuries = sum(INJURIES)) %>% arrange(-total.injuries)
head(df.injuries, 10)

```  

###Economic Impact
The data provides two types of economic impact, namely property damage (PROPDMG) and crop damage (CROPDMG). The actual damage in $USD is indicated by PROPDMGEXP and CROPDMGEXP parameters. According to this link, the index in the PROPDMGEXP and CROPDMGEXP can be interpreted as the following:-  
H, h -> hundreds = x100  
K, K -> kilos = x1,000  
M, m -> millions = x1,000,000  
B,b -> billions = x1,000,000,000  
(+) -> x1  
(-) -> x0  
(?) -> x0  
blank -> x0  
The total damage caused by each event type is calculated with the following code.  

```{r EconomicImpact}
df.damage <- df %>% select(EVTYPE, PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)

Symbol <- sort(unique(as.character(df.damage$PROPDMGEXP)))
Multiplier <- c(0,0,0,1,10,10,10,10,10,10,10,10,10,10^9,10^2,10^2,10^3,10^6,10^6)
convert.Multiplier <- data.frame(Symbol, Multiplier)

df.damage$Prop.Multiplier <- convert.Multiplier$Multiplier[match(df.damage$PROPDMGEXP, convert.Multiplier$Symbol)]
df.damage$Crop.Multiplier <- convert.Multiplier$Multiplier[match(df.damage$CROPDMGEXP, convert.Multiplier$Symbol)]

df.damage <- df.damage %>% mutate(PROPDMG = PROPDMG*Prop.Multiplier) %>% mutate(CROPDMG = CROPDMG*Crop.Multiplier) %>% mutate(TOTAL.DMG = PROPDMG+CROPDMG)

df.damage.total <- df.damage %>% group_by(EVTYPE) %>% summarize(TOTAL.DMG.EVTYPE = sum(TOTAL.DMG))%>% arrange(-TOTAL.DMG.EVTYPE) 

head(df.damage.total,10)
```  
###Results  

##Health Impact  

The top 10 events with the highest total fatalities and injuries are shown graphically.  
```{r PlotPrinting}
library(ggplot2)
g <- ggplot(df.fatalities[1:10,], aes(x=reorder(EVTYPE, -total.fatalities), y=total.fatalities))+geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))+ggtitle("Top 10 Events with Highest Total Fatalities") +labs(x="EVENT TYPE", y="Total Fatalities")
g

g <- ggplot(df.injuries[1:10,], aes(x=reorder(EVTYPE, -total.injuries), y=total.injuries))+geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))+ggtitle("Top 10 Events with Highest Total Injuries") +labs(x="EVENT TYPE", y="Total Injuries")
g

```  

As shown in the figures, tornado causes the hightest in both the total fatality and injury count.  

###Economic Impact  

The top 10 events with the highest total economic damages (property and crop combined) are shown graphically.

```{r ECOImpact}
g <- ggplot(df.damage.total[1:10,], aes(x=reorder(EVTYPE, -TOTAL.DMG.EVTYPE), y=TOTAL.DMG.EVTYPE))+geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))+ggtitle("Top 10 Events with Highest Economic Impact") +labs(x="EVENT TYPE", y="Total Economic Impact ($USD)")

g
```  

As shown in the figure, flood has the highest economic impact.  

